\hspace{1cm} Do syntactic and semantic similarity lead to interference effects?\newline Evidence from self-paced reading and event-related potentials using German

### 2023-10-17 paper/main.tex
added dois

\appendix

\section{Previous work on syntactic and semantic interference in reading}

But are both syntactic and semantic cues used during memory retrieval? \citet{vandyke07} and \citet{mertzen} investigated the effects of syntactic and semantic interference during long-distance subject-verb dependencies (see Example \ref{ex:1_vandyke07}) in a series of self-paced reading and eye-tracking while reading experiments. Overall, the original study by \citet{vandyke07} and the replication by \citet{mertzen} showed relatively consistent results. They consistently found syntactic interference but only rarely semantic interference (only in some measures of some of the experiments and sometimes in regions outside of the critical region). These results are not in line with the predictions from the standard cue-based retrieval models \citep{Lewis2005, mcelree}: While both previous studies found effects of syntactic and semantic interference, they emerged in different regions -- sometimes even prior to the retrieval site. Figure \ref{fig:previous} gives an overview of the previous findings. See the appendix for a detailed discussion of the previous results.

One possibly problematic aspect of the studies from \citet{vandyke07} and \citet{mertzen} is that effects were found in pre-critical regions. \citet{mertzen} suggested that the pre-critical effects could be due to structural confounds, encoding interference, preview effects during natural reading in the eye-tracking experiments or anticipatory retrieval at the pre-critical word because participants could predict that the verb would follow. It is unclear from the previous studies when exactly the pre-critical effects emerged, i.e., when the conditions started to differ. But it is important to know the timecourse of the pre-critical effects to determine their cause. If the structural confounds explanation is correct, then the pre-critical effects should appear only in relation to the syntactic manipulation. For the encoding interference explanation to be correct, it would be necessary that conditions start to differ at (or shortly after) the distractor. The preview and anticipatory retrieval explanations can only apply if the pre-critical effects are found only close to the critical verb.

The previous estimates had relatively wide 95\,\% confidence/credible intervals in some of the experiments, which is probably due to the relatively small sample sizes. \citet{vandyke07} had a typical number of participants and items for psycholinguistic experiments (35 - 40 participants per experiment, 36 - 48 items per experiment). \citet{mertzen} used the same number of items as \citet{vandyke07}, but larger participant sample sizes (English: N = 61, German N = 121). Large uncertainty raises the possibility of Type M errors, leading to misleading conclusions \citep{gelman_carlin}.

\begin{figure}[H]
    \caption{Overview of previous findings of syntactic and semantic interference and their interaction by \citet{vandyke07} (three experiments in English with 35, 36 or 40 participants) and \citet{mertzen} (one experiment in German with 121 participants, one experiment in English with 61 participants). Experiments used either Self-paced Reading (SPR) or Eye-tracking methodology (Eye-Tracking measures: FPRT = first-pass reading time, RPD = regression-path duration, TFT = total fixation time). Error bars are 95\% confidence intervals for \citet{vandyke07} and 95\% credible intervals for \citet{mertzen}. }
    \label{fig:previous}
    \centering
    \includegraphics[width=0.95\textwidth]{images/previous_findings_allregions_long.png}
    \figurenote{The estimates for \citeauthor{vandyke07}'s (\citeyear{vandyke07}) data were derived form the published paper. For non-significant effects, where no statistics were reported, we used a t-value of 1.95 to compute the confidence intervals. \citet{mertzen} provided their estimates and credible intervals.}
\end{figure}

The results of the previous studies raise two questions: 1) Why were semantic interference effects found less consistently than syntactic interference effects? 2) Why were effects of retrieval interference observed in the pre-critical region? To answer question 1, we decided to run large-sample studies revisiting \citeauthor{vandyke07}'s (\citeyear{vandyke07}) design to gain more precise estimates of the effects of syntactic and semantic interference. To answer question 2 -- or to disentangle the possible causes of the pre-critical effects -- we decided it is best to use methods which provide responses which are unambiguously related to the processing of single words. The methods of our choice being self-paced reading (SPR) and electroencephalography (EEG) or event-related potentials (ERPs), respectively, with typical rapid serial visual presentation (RSVP) of single words. Both these methods provide responses directly linked to the processing of single words and, unlike eye-tracking while reading, do not grant any preview of upcoming stimuli. To reduce the structural confound at the pre-critical region identified by \citet{mertzen}, we added an additional pre-pre-critical region. These changes in the experimental design between the present and previous studies will help to distinguish between the different explanations for the pre-critical effects put forward by \citet{mertzen}.

In the previous studies, surprising effects were not only found in the pre-critical region, but also in the sentence-final region: Specifically, \citet{vandyke07} found semantic interference effects at the sentence-final region. She proposed that the effect was delayed to this late region because ``it simply takes longer for individuals to compute the semantic association between the distractor and the verb'' \citep[][p. 427]{vandyke07}. An alternative explanation for this late effect is that eye-movement control constraints can push the eye to move on before processing is finished. In contrast, ERPs are a time-sensitive direct measure of brain activity in response to the stimuli without confounds of planning and execution of motor activity (i.e., planning of saccades during natural reading in eye-tracking experiments or planning of button presses in SPR experiments). The well-studied N400 ERP component is associated with integration in the sentence context and / or memory retrieval induced by the current stimulus \citep{brouwer2017_n4_p6, lau2008_n400, kutas&federmeier_2000, kutas_federmeier2011}. The N400 is a negativity that peaks 400 ms after a word appeared on the screen in word-by-word presentation. The canonical finding is that the N400 amplitude is more negative for words which provide a poor semantic fit into the sentence context compared to words which fit the context perfectly \citep{kutas_hillyard}. Therefore, the N400 can provide insights into the immediate semantic processing of a word. This should prevent a potential semantic interference effect to be only detected at later regions in the sentence.

Although the N400 is primarily associated with semantic processing, if one follows the memory retrieval account it is reasonable to assume that retrieval of all aspects of a memory entry will modulate the N400 amplitude. The potential complications during memory retrieval due to interference should not differ dependent on the specific cue used for retrieval, i.e., \{$\pm$ grammatical subject\} or \{$\pm$ animate\}.  Moreover, previous ERP studies investigating interference have found negativities in the typical N400 time window (300 - 500 ms post-stimulus onset), although the scalp topographies were not typical for the N400, being either broad \citep{schoknecht2022} or left anterior \citep{vasishth_drenhaus_2011}. Therefore, ERPs and especially the N400, are suited to be used for the investigation of interference (in the design from \citet{vandyke07} and in general).  We chose to focus our EPR analyses on the N400 and expected to find interference effects -- both syntactic and semantic -- in N400 amplitude differences due to a more complex memory retrieval under high compared to low interference.

\section{I don't know why this is here}
In their \citeyear{vandyke07} paper, \citeauthor{vandyke07} reported a series of reading studies (one self-paced reading experiment and two eye-tracking experiments). For the eye-tracking experiments, \citet{vandyke07} reported three reading time measures: First-pass reading times (FPRT), regression-path durations (RPD) and total fixaton times (TFT). Reading times in multiple regions were reported, but the critical region was the verb. At the verb, \citet{vandyke07} found syntactic interference in both eye-tracking experiments and in all measures, but semantic interference was only found in Experiment 2 and only in RPD and TFT. In contrast, the spill-over region showed semantic interference more consistently than semantic interference. Surprisingly, the pre-critical region, which was only present in \citeauthor{vandyke07}'s (\citeyear{vandyke07}) Experiment 3, showed main effects of syntactic and semantic interference. On top of that, the RPDs of the sentence-final region in Experiment 2 and 3 were consistent with effects of syntactic and  semantic interference.

\citeauthor{mertzen}'s (\citeyear{mertzen}) estimates at the pre-critical region showed slowdowns for high vs.\ low  syntactic and semantic interference in English and German in RPD and TFT. In the reading times of the critical verb, \citeauthor{mertzen}'s (\citeyear{mertzen}) estimates were consistent with syntactic interference effects in all measures in English and German. Additionally, the estimates suggested a semantic interference effect in English. In the spill-over region, the estimates only suggested a semantic interference effect in German.

>>>>

%\appendix

%\section{Previous work on syntactic and semantic interference in reading}

%Are both syntactic and semantic cues used during memory retrieval? \citet{vandyke07} and \citet{mertzen} investigated the effects of syntactic and semantic interference during long-distance subject-verb dependencies (see Example \ref{ex:1_vandyke07}) in a series of self-paced reading and eye-tracking while reading experiments. Overall, the original study by \citet{vandyke07} and the replication by \citet{mertzen} showed relatively consistent results. They consistently found syntactic interference but only rarely semantic interference (only in some measures of some of the experiments and sometimes in regions outside of the critical region). These results are not in line with the predictions from the standard cue-based retrieval models \citep{Lewis2005, mcelree}: While both previous studies found effects of syntactic and semantic interference, they emerged in different regions -- sometimes even prior to the retrieval site. Figure \ref{fig:previous} gives an overview of the previous findings.

%One possibly problematic aspect of the studies from \citet{vandyke07} and \citet{mertzen} is that effects were found in pre-critical regions. \citet{mertzen} suggested that the pre-critical effects could be due to structural confounds, encoding interference, preview effects during natural reading in the eye-tracking experiments or anticipatory retrieval at the pre-critical word because participants could predict that the verb would follow. It is unclear from the previous studies when exactly the pre-critical effects emerged, i.e., when the conditions started to differ. But it is important to know the timecourse of the pre-critical effects to determine their cause. If the structural confounds explanation is correct, then the pre-critical effects should appear only in relation to the syntactic manipulation. For the encoding interference explanation to be correct, it would be necessary that conditions start to differ at (or shortly after) the distractor. The preview and anticipatory retrieval explanations can only apply if the pre-critical effects are found only close to the critical verb.

%The previous estimates had relatively wide 95\,\% confidence/credible intervals in some of the experiments, which is probably due to the relatively small sample sizes. \citet{vandyke07} had a typical number of participants and items for psycholinguistic experiments (35 - 40 participants per experiment, 36 - 48 items per experiment). \citet{mertzen} used the same number of items as \citet{vandyke07}, but larger participant sample sizes (English: N = 61, German N = 121). Large uncertainty raises the possibility of Type M errors, leading to misleading conclusions \citep{gelman_carlin}.

%\begin{figure}[H]
%    \caption{Overview of previous findings of syntactic and semantic interference and their interaction by \citet{vandyke07} (three experiments in English with 35, 36 or 40 participants) and \citet{mertzen} (one experiment in German with 121 participants, one experiment in English with 61 participants). Experiments used either Self-paced Reading (SPR) or Eye-tracking methodology (Eye-Tracking measures: FPRT = first-pass reading time, RPD = regression-path duration, TFT = total fixation time). Error bars are 95\% confidence intervals for \citet{vandyke07} and 95\% credible intervals for \citet{mertzen}. }
%    \label{fig:previous}
%    \centering
%    \includegraphics[width=0.95\textwidth]{images/previous_findings_allregions_long.png}
%    \figurenote{The estimates for \citeauthor{vandyke07}'s (\citeyear{vandyke07}) data were derived form the published paper. For non-significant effects, where no statistics were reported, we used a t-value of 1.95 to compute the confidence intervals. \citet{mertzen} provided their estimates and credible intervals.}
%\end{figure}

%The results of the previous studies raise two questions: 1) Why were semantic interference effects found less consistently than syntactic interference effects? 2) Why were effects of retrieval interference observed in the pre-critical region? To answer question 1, we decided to run large-sample studies revisiting \citeauthor{vandyke07}'s (\citeyear{vandyke07}) design to gain more precise estimates of the effects of syntactic and semantic interference. To answer question 2 -- or to disentangle the possible causes of the pre-critical effects -- we decided it is best to use methods which provide responses which are unambiguously related to the processing of single words. The methods of our choice being self-paced reading (SPR) and electroencephalography (EEG) or event-related potentials (ERPs), respectively, with typical rapid serial visual presentation (RSVP) of single words. Both these methods provide responses directly linked to the processing of single words and, unlike eye-tracking while reading, do not grant any preview of upcoming stimuli. To reduce the structural confound at the pre-critical region identified by \citet{mertzen}, we added an additional pre-pre-critical region. These changes in the experimental design between the present and previous studies will help to distinguish between the different explanations for the pre-critical effects put forward by \citet{mertzen}.

%In the previous studies, surprising effects were not only found in the pre-critical region, but also in the sentence-final region: Specifically, \citet{vandyke07} found semantic interference effects at the sentence-final region. She proposed that the effect was delayed to this late region because ``it simply takes longer for individuals to compute the semantic association between the distractor and the verb'' \citep[][p. 427]{vandyke07}. An alternative explanation for this late effect is that eye-movement control constraints can push the eye to move on before processing is finished. In contrast, ERPs are a time-sensitive direct measure of brain activity in response to the stimuli without confounds of planning and execution of motor activity (i.e., planning of saccades during natural reading in eye-tracking experiments or planning of button presses in SPR experiments). The well-studied N400 ERP component is associated with integration in the sentence context and / or memory retrieval induced by the current stimulus \citep{brouwer2017_n4_p6, lau2008_n400, kutas&federmeier_2000, kutas_federmeier2011}. The N400 is a negativity that peaks 400 ms after a word appeared on the screen in word-by-word presentation. The canonical finding is that the N400 amplitude is more negative for words which provide a poor semantic fit into the sentence context compared to words which fit the context perfectly \citep{kutas_hillyard}. Therefore, the N400 can provide insights into the immediate semantic processing of a word. This should prevent a potential semantic interference effect to be only detected at later regions in the sentence.

%Although the N400 is primarily associated with semantic processing, if one follows the memory retrieval account it is reasonable to assume that retrieval of all aspects of a memory entry will modulate the N400 amplitude. The potential complications during memory retrieval due to interference should not differ dependent on the specific cue used for retrieval, i.e., \{$\pm$ grammatical subject\} or \{$\pm$ animate\}.  Moreover, previous ERP studies investigating interference have found negativities in the typical N400 time window (300 - 500 ms post-stimulus onset), although the scalp topographies were not typical for the N400, being either broad \citep{schoknecht2022} or left anterior \citep{vasishth_drenhaus_2011}. Therefore, ERPs and especially the N400, are suited to be used for the investigation of interference (in the design from \citet{vandyke07} and in general).  We chose to focus our EPR analyses on the N400 and expected to find interference effects -- both syntactic and semantic -- in N400 amplitude differences due to a more complex memory retrieval under high compared to low interference.

%\section{I don't know why this is here}
%In their \citeyear{vandyke07} paper, \citeauthor{vandyke07} reported a series of reading studies (one self-paced reading experiment and two eye-tracking experiments). For the eye-tracking experiments, \citet{vandyke07} reported three reading time measures: First-pass reading times (FPRT), regression-path durations (RPD) and total fixaton times (TFT). Reading times in multiple regions were reported, but the critical region was the verb. At the verb, \citet{vandyke07} found syntactic interference in both eye-tracking experiments and in all measures, but semantic interference was only found in Experiment 2 and only in RPD and TFT. In contrast, the spill-over region showed semantic interference more consistently than semantic interference. Surprisingly, the pre-critical region, which was only present in \citeauthor{vandyke07}'s (\citeyear{vandyke07}) Experiment 3, showed main effects of syntactic and semantic interference. On top of that, the RPDs of the sentence-final region in Experiment 2 and 3 were consistent with effects of syntactic and  semantic interference.

%\citeauthor{mertzen}'s (\citeyear{mertzen}) estimates at the pre-critical region showed slowdowns for high vs.\ low  syntactic and semantic interference in English and German in RPD and TFT. In the reading times of the critical verb, \citeauthor{mertzen}'s (\citeyear{mertzen}) estimates were consistent with syntactic interference effects in all measures in English and German. Additionally, the estimates suggested a semantic interference effect in English. In the spill-over region, the estimates only suggested a semantic interference effect in German.
